Review Paper Draft

### 2020-06-06 main.tex
Section 4 updates

\begin{figure}[H]
\centering
\begin{tikzpicture}[%
>=latex',
circ/.style={draw, shape=circle, node distance=1cm, line width=1pt}]%Define the arrow type and style for circled nodes
\draw[dashed] (0,1) node[left] (P) {Population Stratification} -- (2,1) node[right] (E) {Environment}; 
\draw[->] (0,0) node[left] (X) {Observed SNPs} -- (2,0) node[right] (Y) {Phenotype}; %Line between X and Y
\draw[->] (P)   -- (X) ; 
\draw[->] (E)   -- (Y) ; 
% \filldraw[color=red!60, fill=none, very thick] (0,1) ellipse (4.5 and 0.35);
\end{tikzpicture}
\end{figure}

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix.

LMM-Lasso-Rakitsch \cite{Rakitsch2012} uses a two-step procedure for estimating the variance components, in which the variance components and corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \cite{glmnet} or \texttt{ncvreg} \cite{ncvreg}. LMM-Lasso-ggmix \cite{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \cite{ggmix, bhatnagar2019simultaneous}. \anna{We postpone our discussion of the differences in performance of each method to the results section.}

@@

We generate these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian, African American, and Hispanic populations. \anna{cite Vanderwas data somehow, add n = ??? for all three groups).} For each $\bX$ matrix, $\bZ$ is a corresponding incidence matrix of 1s and 0s identifying subpopulation membership. Several structures of the fixed $\bgamma$ were considered. \anna{We give more details about this in the results section.}

@@

%------------------------------%
\subsection{LMM-Lasso outperforms Lasso - \anna{keep low-dimensional results? Only 'real data' for now... could use VanderWas to generate a $y$...}}
%------------------------------%
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/fig1a.png}
    \caption{Low-dimensional setting ($n=200, p=100$)}
    \label{fig:ld}
\end{figure}

%------------------------------%
\subsection{PC-Lasso and LMM-Lasso outperform Lasso}
%------------------------------%
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/fig1b.png}
    \caption{High dimensional setting ($n=200, p=1000$)}
    \label{fig:hd}
\end{figure}

\anna{add high-dimensional real data case using vanderwas?}

%------------------------------%
\subsection{Population structure}
%------------------------------%
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/kinship1}
    \caption{A few large subpopulations}
    \label{fig:big_blocks}
\end{figure}
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/kinship2}
    \caption{Many small subpopulations}
    \label{fig:small_blocks}
\end{figure}

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso when simulated subpopulations are small}
%------------------------------%
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/fig2a}
    \caption{Relative performance of LMM-Lasso, PC-Lasso in terms of estimation accuracy.}
    \label{fig:big_vs_small}
\end{figure}

%------------------------------%
\subsection{$\gamma$ structure}
%------------------------------%
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/gamma_structure.png}
    \caption{$\bgamma$ structures}
    \label{fig:gamma_structures}
\end{figure}

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso when $\gamma$ is heterogeneous}
%------------------------------%
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/fig2b.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for different subpopulation and $\bgamma$ structures}
    \label{fig:my_label}
\end{figure}

%------------------------------%
\subsection{Re-estimation of $\eta$ may, in some cases, slightly improve our estimate of $\hat{\eta}$, but does not improve the estimation of $\beta$, and takes a long time}
Put this at the end of results
%------------------------------%

\anna{We have been unable to find a scenario where ggmix outperforms the LMM-Lasso method implemented by \cite{Rakitsch2012}. Ggmix performs poorly in a low-dimensional setting due to the ill-conditioning of the matrix K. Although it is primarily intended for use in high dimensional settings, where it performs on par with LMM-Lasso, this is also the scenario where it is ~100 times slower than LMM-Lasso. }\\

@@

\anna{ggmix vs lmmlasso prc plot (?)}\\

\anna{ggmix vs. lmmlasso time plot}- I have this, make it pretty\\

>>>>

\begin{figure}[H]
\centering
\begin{tikzpicture}[%
>=latex',
circ/.style={draw, shape=circle, node distance=1cm, line width=1pt}]%Define the arrow type and style for circled nodes
\draw[dashed] (0,1) node[left] (P) {Population Stratification} -- (2,1) node[right] (E) {Environment}; 
\draw[->] (0,0) node[left] (X) {Observed SNPs} -- (2,0) node[right] (Y) {Phenotype}; %Line between X and Y
\draw[->] (P)   -- (X) ; 
\draw[->] (E)   -- (Y) ; 
% \filldraw[color=red!60, fill=none, very thick] (0,1) ellipse (4.5 and 0.35);
\end{tikzpicture}
\caption{Directed acyclic graph (DAG) of confounding due to population structure}
\label{fig:ps_env}
\end{figure}

@@

We have so far used the term $\bW$ to describe the general LMM-Lasso method. However, the form of $\bW$ and method of variance component estimation differs between LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch \cite{Rakitsch2012} uses a two-step procedure for estimating the variance components, in which the variance components and corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \cite{glmnet} or \texttt{ncvreg} \cite{ncvreg}. LMM-Lasso-ggmix \cite{bhatnagar2019simultaneous} proposes iteratively updating $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \cite{ggmix, bhatnagar2019simultaneous}. \anna{We postpone our discussion of the differences in performance of each method to the results section.}

@@

We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian (\anna{n = ?}), African American (\anna{n = ?}), and Hispanic (\anna{n = ?}). \anna{cite Vanderwas data} For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation membership. 500 simulations were performed for each data type and parameter combination.

@@

%------------------------------%
\subsection{PC-Lasso and LMM-Lasso outperform Lasso}
%------------------------------%
Firstly, these penalized regression methods were compared in the presence of varying levels of environmental confounding. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) in various settings and broken down into bias and variance components.  Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of figure \ref{fig:mse} shows that when $\xi = 0$, and no environmental confounding is present, lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that the top 10 principal components do not explain any variability in the outcome, yet are included in the final model, since they are unpenalized. The bottom row of figure \ref{fig:mse} illustrates the resulting MSE when $\xi = 0.8$ and thus 40\% of the variability of the phenotype is attributable to environmental confounding. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpoplulations simulated data due to the fact that the subpopulations are more easily distinguishable based on their observed genetic data. \anna{need details about the real data once it's run}.\\

\anna{Do I need to discuss the large subpops, heterogeneous gamma here?} \\

\anna{add high-dimensional real data case using vanderwas? add dashed line to legend = null model mse}\\
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/fig1b.png}
    \caption{MSE for different data types in the presence and absence of environmental confounding.}
    \label{fig:mse}
\end{figure}

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso when simulated subpopulations are small}
%------------------------------%
Given the superior performance of PC-Lasso and LMM-Lasso compared to the naive lasso in the presence of environmental confounding, we next compared these methods in the presence of two forms of simulated population structure: (1) relatedness characterized by a small number of large subpopulations \anna{with moderate relatedness (?)} to model large-scale environmental heterogeneity or population stratification, similar to that described in our motivating air qualitty example, and (2) a large number of small, closely related subpopulations intended to approximate finer grained environmental heterogeneity, such as household exposures or behavioral norms. Figure \ref{fig:kinship} illustrates the simulated kinship structures in these scenarios, \anna{as well as that present in the empirical SNP data}. We note that we show results for simulated data generated such that subpopulations were of equal size. Those of unequal size were also considered, but did not result in substantively different results (not shown - could be in supplement).

\begin{figure}[H]
\begin{subfigure}{.8\textwidth}
  \centering
  % include first image
  \includegraphics[width=.7\linewidth]{figures/kinship1}  
  \caption{Independent subpopulations simulated kinship; a small number of large groups}
  \label{fig:large_groups}
\end{subfigure}
\newline
\begin{subfigure}{.8\textwidth}
  \centering
  % include second image
  \includegraphics[width=.7\linewidth]{figures/kinship2}  
  \caption{Independent subpopulations simulated kinship; a large number of small groups}
  \label{fig:small_groups}
\end{subfigure}
\newline
\begin{subfigure}{.8\textwidth}
  \centering
  % include second image
  \includegraphics[width=.7\linewidth]{figures/vanderwas_kinship.pdf}  
  \caption{\anna{Vanderbilt data genome-wide} kinship}
  \label{fig:vandy_groups}
\end{subfigure}
\caption{\anna{obviously these figures need to be tweaked:sizes, uniform scales} \\ \anna{Should kinship in the first figure be lower - more moderate?}}
\label{fig:kinship}
\end{figure}

Pseudophenotypes for these simulations were again generated using a 1:1 SNR, $\eta = 0.5$, and in the presence of moderate environmental confounding, $\xi = 0.5$. \anna{Due to constraints within the BNPSD package and the nature of the admixed population model they use (???), only the independent subpopulations method was able to simulate a large number of small subpopulations. For this reason, the presented simulated results all correspond to this simulation method from the bnpsd package.} Figure \ref{fig:big_vs_small} summarizes the results of these simulations with the y-axis displaying the difference in MSE (PC-Lasso - LMM-Lasso) for each simulation and for the various subpopulation structures. The dashed horizontal line at 0 indicates no difference in the performance of PC-Lasso and LMM-Lasso, while differences greater than 0 correspond to smaller MSE for LMM-Lasso, and thus superior estimation accuracy. In the presence of a small number of large subpopulations, there is some overlap in the estimation accuracy of PC-Lasso and LMM-Lasso, as evidenced by the corresponding box straddling the 0 line. However, the majority of the box falls below 0, indicating the superior performance of PC-Lasso in many of the simulations. IN the presence of many small subpopulations, however, the entire box and error bars remain above the 0 line, indicating the clear superior estimation accuracy of LMM-Lasso in this context. \cite{hoffman2013correcting} showed that principal component adjustment methods represent a low dimensional approximation to linear mixed models. As such, it makes sense that PC-Lasso, which corrects for population structure using a low dimensional set of surrogate variables, would perform well in the large subpopulation setting, but fall short in the presence of many small subgroups. \anna{Additionally, its performance in this context may also me augmented by the fact that the subpopulations are more discretized  easily distinguishable based on their observed genetic data in the independent subpopulation simulated data. Although small subpopulations are unable to be simulated in the BNPSD admixed subpopulation setting, the corresponding results for the large subgroups in the admixed simulated data show much more comparable results between PC-Lasso and LMM-Lasso. - do this}. \anna{Vanderwas?}

\anna{Should Vanderwas simulated data be here?}
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.7]{figures/fig2a}
    \caption{Relative performance of LMM-Lasso, PC-Lasso in terms of estimation accuracy.}
    \label{fig:big_vs_small}
\end{figure}

%------------------------------%
\subsection{LMM-Lasso outperforms PC-Lasso when $\gamma$ is heterogeneous}
%------------------------------%

In addition to considering the effect of population structure, or forms of $\bZ$, on these methods' estimation accuracy, we also evaluated the effect of varying forms of $\bgamma$, the subpopulation-specific effect sizes. Our initial thinking was that structures of $\bgamma$ that can be well-approximated by a linear combination of the observed genetic data would be well-approximated and corrected for by PC-Lasso, whereas LMM-Lasso may be more able to correct for subpopulation-specific effects that do not lend themselves to linear approximations. To that end we investigated four distinct $\bgamma$ structures: linear homogeneous, linear heterogeneous, nonlinear homogeneous, and nonlinear heterogeneous, \anna{names should change... half and half, not exponential is the better nonlinear setting}. In our naming convention, linear/nonlinear refers to whether the coefficients follow a linear pattern versus a nonlinear form, while homogeneous/heterogeneous indicates whether coefficients have the same, or differing signs, respectively. For example, (1, -2, 3, -4) is an example of a linear-heterogeneous structure, whereas (2, 2, 1, 1) is a nonlinear-homogeneous form. Figure \ref{fig:gamma_structures} displays a visual summary of these coefficient structures. \\

\anna{put half and half in this plot instead of exponential? In addition?}
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{$\bgamma$ structures}
    \label{fig:gamma_structures}
\end{figure}

We thus evaluate PC-Lasso and LMM-Lasso in the presence of large and small subpopulations as well as distinct coefficient structures. Figure \ref{fig:big_small_gamma} shows the difference in MSE (PC-Lasso - LMM-Lasso) for all simulations, broken down by both subpopulation and coefficient structure. The right-hand panels of \ref{fig:big_small_gamma} are consistent with figure \ref{fig:big_vs_small}, showing superior LMM-Lasso performance in the presence of small subpopulations, regardless of coefficient linearity or homogeneity. The panel of figure \ref{fig:big_small_gamma} corresponding to large subpopulations with heterogeneous coefficients also appears to be a setting where LMM-Lasso consistently outperforms PC-Lasso, regardless of coefficient linearity. Only in the setting of large supopulations with homogeneous coefficients does PC-Lasso outperform LMM-Lasso. In line with our initial intuition, PC-Lasso performs best within this setting when coefficients are linear in nature.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/fig2b.png}
    \caption{Relative performance of LMM-Lasso, PC-Lasso for different subpopulation and $\bgamma$ structures}
    \label{fig:big_small_gamma}
\end{figure}

%------------------------------%
\subsection{Re-estimation of $\eta$ may, in some cases, slightly improve our estimate of $\hat{\eta}$, but does not improve the estimation of $\beta$, and takes a long time}
%------------------------------%

In section 3 briefly mentioned the existence of to LMM-Lasso variants: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. The LMM-Lasso results shown so far are those of LMM-Lasso-Rakitsch, though they are representative of LMM-Lasso-ggmix. The primary purpose of this review is to compare LMM-Lasso to other penalized regression methods which aim to correct for population structure and environmental heterogeneity. However, we also want to explain why we used LMM-Lasso-Rakitsch, and provide guidance for potential users of either method.

LMM-Lasso-Rakitsch estimates the variance components once, under the assumption of a null model. LMM-Lasso-ggmix seeks to improve upon this by re-estimating the variance components and $\bbeta$ iteratively. Although this underlying idea seems reasonable, we have been unable to find a scenario where LMM-Lasso-ggmix outperforms the LMM-Lasso-Rakitsch in estimation accuracy of $\bbeta$ or $\eta$.

@@

In addition, we have found that LMM-Lasso-ggmix performs poorly in a low-dimensional setting due to the ill-conditioning of the RRM. Given its intended use for the analysis of high-dimensional data, this fact could be of little consequence to practitioners. However, although its estimation accuracy in high dimensions is on par with that of LMM-Lasso-Rakitsch, \anna{as shown in two previous mse figs}, this is also a setting in which the computational time required may be prohibitive for practical applications. Figure \anna{time fig} shows a ratio of the computing time required for LMM-Lasso-ggmix compared to that of LMM-Lasso-Rakitsch, where the horizontal dashed line at 1 indicates both methods require the same computing time. In high dimenions, where LMM-Lasso-ggmix performs well, it is between \anna{something and something else} times slower than LMM-Lasso-Rakitsch. Thus, LMM-Lasso-Rakitsch seems to be a more attractive and scalable method. We have implemented the LMM-Lasso-Rakitsch method using the equivalent but more interpretable $\eta$ paramaterization in the \texttt{R} package, \texttt{penalizedLMM - or something more fun}, available on the author's github page: \url{https://github.com/areisett/penalizedLMM}.
